---
title: "Preprocessing"
author: "Linfeng Hu"
date: "2024-02-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Datasets
```{r}
library(purrr)
library(forcats)
library(haven)
library(dplyr)
library(tidyr)

folder_path <- "/Users/linfenghu/Desktop/SAMHSA_rdata"
rdata_files <- list.files(folder_path, pattern = "\\.rdata$", full.names = TRUE)
for (file in rdata_files) {load(file)}
```

## Aggregate based on STATE

```{r}
state_map <- c(
  "1" = "Alabama", "2" = "Alaska", "4" = "Arizona", "5" = "Arkansas", "6" = "California", "8" = "Colorado",
  "9" = "Connecticut", "10" = "Delaware", "11" = "District of Columbia", "12" = "Florida",
  "13" = "Georgia", "15" = "Hawaii", "39" = "Ohio", "23" = "Maine", "42" = "Pennsylvania", "46" = "South Dakota",
  "16" = "Idaho", "17" = "Illinois", "18" = "Indiana", "19" = "Iowa", "20" = "Kansas", "21" = "Kentucky",
  "22" = "Louisiana", "24" = "Maryland", "25" = "Massachusetts", "26" = "Michigan", "27" = "Minnesota",
  "28" = "Mississippi", "29" = "Missouri", "30" = "Montana", "31" = "Nebraska", "32" = "Nevada",
  "33" = "New Hampshire", "34" = "New Jersey", "35" = "New Mexico", "36" = "New York", "37" = "North Carolina",
  "38" = "North Dakota", "40" = "Oklahoma", "41" = "Oregon", "44" = "Rhode Island", "45" = "South Carolina",
  "47" = "Tennessee", "48" = "Texas", "49" = "Utah", "50" = "Vermont", "51" = "Virginia", "53" = "Washington",
  "54" = "West Virginia", "55" = "Wisconsin", "56" = "Wyoming", "72" = "Puerto Rico", "99" = "Other Jurisdiction"
)
```

```{r}
aggregate_dataset <- function(dataset) {
  labelled_columns <- names(dataset)[sapply(dataset, is.labelled)]
  dataset[labelled_columns] <- lapply(dataset[labelled_columns], as.integer)
  
  aggregated_data <- dataset %>%
    group_by(STATEFIP) %>%
    summarize(
      YEAR = first(YEAR),  # Keep the first value of YEAR
      POPULATION = n(),
      SPHSERVICE = sum(SPHSERVICE == 1, na.rm = TRUE),  
      CMPSERVICE = sum(CMPSERVICE == 1, na.rm = TRUE),
      OPISERVICE = sum(OPISERVICE == 1, na.rm = TRUE),  
      RTCSERVICE = sum(RTCSERVICE == 1, na.rm = TRUE),  
      IJSSERVICE = sum(IJSSERVICE == 1, na.rm = TRUE),  
      MH1_values = list(table(factor(MH1, levels = c(-9,1:13))))
    ) %>%
    ungroup() %>%
    mutate(MH1_values = map(MH1_values, ~setNames(as.list(.x), paste0(names(.x))))) %>%
    unnest_wider(MH1_values, names_sep = "_") %>%
    replace(is.na(.), 0)  # Replace NA values with 0
  return(aggregated_data)
}
```

```{r}
# Apply the aggregation function to each dataset
df_2013 <- aggregate_dataset(mhcld_puf_2013_r)
df_2014 <- aggregate_dataset(mhcld_puf_2014_r)
df_2015 <- aggregate_dataset(mhcld_puf_2015_r)
df_2016 <- aggregate_dataset(mhcld_puf_2016_r)
df_2017 <- aggregate_dataset(mhcld_puf_2017_r)
df_2018 <- aggregate_dataset(mhcld_puf_2018_r)
df_2019 <- aggregate_dataset(mhcld_puf_2019_r)
df_2020 <- aggregate_dataset(mhcld_puf_2020_r)
df_2021 <- aggregate_dataset(mhcld_puf_2021_r)

df_2013$STATEFIP <- state_map[as.character(df_2013$STATEFIP)]
df_2014$STATEFIP <- state_map[as.character(df_2014$STATEFIP)]
df_2015$STATEFIP <- state_map[as.character(df_2015$STATEFIP)]
df_2016$STATEFIP <- state_map[as.character(df_2016$STATEFIP)]
df_2017$STATEFIP <- state_map[as.character(df_2017$STATEFIP)]
df_2018$STATEFIP <- state_map[as.character(df_2018$STATEFIP)]
df_2019$STATEFIP <- state_map[as.character(df_2019$STATEFIP)]
df_2020$STATEFIP <- state_map[as.character(df_2020$STATEFIP)]
df_2021$STATEFIP <- state_map[as.character(df_2021$STATEFIP)]
```

```{r}
df_list <- list(df_2013, df_2014, df_2015, df_2016, df_2017, df_2018, df_2019, df_2020, df_2021)
stacked_df <- bind_rows(df_list)
write.csv(stacked_df, "stacked_data_state.csv", row.names = FALSE)
```

diag_map <- c(
  "1" = "Trauma/Stress-related Disorder", "2" = "Anxiety Disorder", "3" = "ADHD", "4" = "Conduct Disorder", 
  "5" = "Delirium/Dementia","6" = "Bipolar Disorder", "7" = "Depressive Disorder", 
  "8" = "Oppositional Defiant Disorder", "9" = "Pervasive Developmental Disorders" ,
  "10" = "Personality Disorders", "11" = "Schizophrenia or other Psychotic Disorders",
  "12" = "Alcohol/Substance Use Disorders", "13" = "Other Disorder/Condition", "-9" = "Missing"
)

```{r}
df <- read.csv("stacked_data_state.csv")
df <- df %>%
  rename(
    "Trauma/Stress-related Disorder" = "MH1_values_1", "Anxiety Disorder" = "MH1_values_2", 
    "ADHD" = "MH1_values_3", "Conduct Disorder" = "MH1_values_4", "Delirium/Dementia" = "MH1_values_5",
    "Bipolar Disorder" = "MH1_values_6", "Depressive Disorder" = "MH1_values_7", 
    "Oppositional Defiant Disorder" = "MH1_values_8", "Pervasive Developmental Disorders" = "MH1_values_9", 
    "Personality Disorders" = "MH1_values_10", "Schizophrenia or other Psychotic Disorders" = "MH1_values_11",
    "Alcohol/Substance Use Disorders" = "MH1_values_12", "Other Disorder/Condition" = "MH1_values_13",
    "Missing" = "MH1_values_.9"
  )
write.csv(df, "stacked_data_state.csv", row.names = FALSE)
```


## Merge all dataframes based on STATE
```{r}
add_year_to_columns <- function(df, year) {
  rename_cols <- setdiff(names(df), c("YEAR", "STATEFIP"))
  new_names <- paste0(rename_cols, "_", year)
  names(df)[names(df) %in% rename_cols] <- new_names
  return(df)
}

remove_year_column <- function(df) {
  df <- df[, !names(df) %in% c("YEAR")]
  return(df)
}
df_list <- list(df_2013, df_2014, df_2015, df_2016, df_2017, df_2018, df_2019, df_2020, df_2021)
years <- c(2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021)
for (i in seq_along(df_list)) {
  df_list[[i]] <- add_year_to_columns(df_list[[i]], years[i])
  df_list[[i]] <- remove_year_column(df_list[[i]])
}
merged_df <- Reduce(function(x, y) merge(x, y, by = "STATEFIP", all = TRUE), df_list)
write.csv(merged_df, "merged_data_state.csv", row.names = FALSE)
```




## Aggregate based on DIAGNOSIS
```{r}
aggregate_dataset_Diagnosis <- function(dataset) {
  result <- dataset %>%
    group_by(YEAR, MH1) %>%
    summarise(
      Male = mean(GENDER == 1, na.rm = TRUE),
      SpecialEdu = mean(EDUC == 1, na.rm = TRUE),
      Edu8 = mean(EDUC == 2, na.rm = TRUE),
      Edu12 = mean(EDUC == 3 | EDUC == 4, na.rm = TRUE),
      EduHigh = mean(EDUC == 5, na.rm = TRUE),
      Black = mean(RACE == 3, na.rm = TRUE),
      White = mean(RACE == 5, na.rm = TRUE),
      OtherRace = mean(RACE %in% c(1,2,4,6), na.rm = TRUE),
      NeverMarried = mean(MARSTAT == 1, na.rm = TRUE),
      NeverMarried = mean(MARSTAT == 1, na.rm = TRUE),
      MarriageHistory = mean(MARSTAT %in% c(2,3,4), na.rm = TRUE),
      NotSMISED = mean(SMISED == 3, na.rm = TRUE),
      SAP = mean(SAP == 1, na.rm = TRUE),
      PrivateResidence = mean(LIVARAG == 2, na.rm = TRUE)
      )
}
```

```{r}
df_2013 <- aggregate_dataset_Diagnosis(mhcld_puf_2013_r)
df_2014 <- aggregate_dataset_Diagnosis(mhcld_puf_2014_r)
df_2015 <- aggregate_dataset_Diagnosis(mhcld_puf_2015_r)
df_2016 <- aggregate_dataset_Diagnosis(mhcld_puf_2016_r)
df_2017 <- aggregate_dataset_Diagnosis(mhcld_puf_2017_r)
df_2018 <- aggregate_dataset_Diagnosis(mhcld_puf_2018_r)
df_2019 <- aggregate_dataset_Diagnosis(mhcld_puf_2019_r)
df_2020 <- aggregate_dataset_Diagnosis(mhcld_puf_2020_r)
df_2021 <- aggregate_dataset_Diagnosis(mhcld_puf_2021_r)
```

```{r}
diag_map <- c(
  "1" = "Trauma/Stress-related Disorder", "2" = "Anxiety Disorder", "3" = "ADHD", "4" = "Conduct Disorder", 
  "5" = "Delirium/Dementia","6" = "Bipolar Disorder", "7" = "Depressive Disorder", 
  "8" = "Oppositional Defiant Disorder", "9" = "Pervasive Developmental Disorders" ,
  "10" = "Personality Disorders", "11" = "Schizophrenia or other Psychotic Disorders",
  "12" = "Alcohol/Substance Use Disorders", "13" = "Other Disorder/Condition", "-9" = "Missing"
)
```

```{r}
df_2013$MH1 <- diag_map[as.character(df_2013$MH1)]
df_2014$MH1 <- diag_map[as.character(df_2014$MH1)]
df_2015$MH1 <- diag_map[as.character(df_2015$MH1)]
df_2016$MH1 <- diag_map[as.character(df_2016$MH1)]
df_2017$MH1 <- diag_map[as.character(df_2017$MH1)]
df_2018$MH1 <- diag_map[as.character(df_2018$MH1)]
df_2019$MH1 <- diag_map[as.character(df_2019$MH1)]
df_2020$MH1 <- diag_map[as.character(df_2020$MH1)]
df_2021$MH1 <- diag_map[as.character(df_2021$MH1)]
```


```{r}
df_list <- list(df_2013, df_2014, df_2015, df_2016, df_2017, df_2018, df_2019, df_2020, df_2021)
stacked_df <- bind_rows(df_list)
write.csv(stacked_df, "merged_data_diagnosis.csv", row.names = FALSE)
```

